```{r setup_ch13, echo = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  out.width = "1000px"
)
```

```{r, echo = FALSE}
library(JuliaCall)
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.9.app/Contents/Resources/julia/bin")
```

```{julia}
using FundamentalsNumericalComputation
default(legend = false)
```

# Two-dimensional problems

**Learning objectives:**

- Expressing functions on a grid
- Reshaping matrices into linear systems
- Poisson (Laplace) equations as a linear system


## Tensor-product discretizations {-}

- We extend one-dimensional problem to the two-dimensional rectangle
- Tensor-product grid:

$$
\{(x_i, y_j): i = 0,...,m; j = 0,...,n\}.
$$

- Functions on the grid are a matrix with elements of $f(x_i,y_j)$

```{julia}
m = 4;   x = range(0,2,length=m+1);
n = 2;   y = range(1,3,length=n+1);
f = (x,y) -> cos(π*x*y-y);
F = [ f(x,y) for x in x, y in y ];
plotlyjs();  # use better 3D renderer
m = 60;   x = range(0,2,length=m+1);
n = 50;   y = range(1,3,length=n+1);
F = [ f(x,y) for x in x, y in y ];

plot(x,y,F',levels=10,fill=true,aspect_ratio=1,
    color=:redsblues,clims=(-1,1),
    xlabel="x",ylabel="y")
surface(x,y,F',l=0,leg=:none,
    color=:redsblues,clims=(-1,1),
    xlabel="x",ylabel="y",zlabel="f(x,y)")
```

- Parametrized surfaces allow us to define non-rectangular shapes, i.e., polar coordinates

```{julia}
r = range(0,1,length=41);
theta = range(0,2π,length=81);
F = [ 1-r^4 for r in r, theta in theta ];

surface(r,theta,F',legend=:none,l=0,color=:viridis,
    xlabel="r",ylabel="theta",title="A polar function")
    
X = [ r*cos(theta) for r in r, theta in theta ];
Y = [ r*sin(theta) for r in r, theta in theta ];

surface(X',Y',F',legend=:none,l=0,color=:viridis,
    xlabel="x",ylabel="y",title="Function on the unit disk")
```

- Partial derivatives for collocation involve transposing for y dimension:

$$
\text{mtx}\left(\frac{\delta u}{\delta x} \right) \approx \mathbf{D}_x \text{mtx}(u), \quad \text{mtx}\left(\frac{\delta u}{\delta y} \right) \approx \text{mtx}(u) \mathbf{D}_y^T 
$$

## Exercises {-}

1a. Make side-by-side surface and contour plots of $f(x,y) = 2y + e^{x-y}, [0,2] \times [-1,1]$

```{julia}
f = (x,y) -> 2*y + exp(x-y);
plotlyjs();  # use better 3D renderer
m = 60;   x = range(0,2,length=m+1);
n = 50;   y = range(-1,1,length=n+1);
F = [ f(x,y) for x in x, y in y ];

plot(x,y,F', layout=(1,2), levels=10,fill=true,aspect_ratio=1,
    color=:redsblues,
    xlabel="x",ylabel="y");
surface!(x,y,F', l=0,leg=:none,
    subplot=2, color=:redsblues,
    xlabel="x",ylabel="y",zlabel="f(x,y)")
```

2a. make side-by-side surface plots of $f_x$ and $f_y$ using Chebyshev spectral differentiation matrix

```{julia}
_,Dx = FNC.diffcheb(m,[0,2]);
_,Dy = FNC.diffcheb(n,[-1,1]);

surface(x,y,(Dx*F)',layout=(1,2), l=0,leg=:none,
    color=:redsblues,
    xlabel="x",ylabel="y",zlabel="df(x,y)/dx");
surface!(x,y,(F*Dy')', l=0,leg=:none,
    subplot=2, color=:redsblues,
    xlabel="x",ylabel="y",zlabel="df(x,y)/dy")
```

3a. Make a contour plot of the mixed derivative $f_{xy}$

```{julia}
plot(x,y,(Dx*F*Dy')', levels=10,fill=true,aspect_ratio=1,
    color=:redsblues, clims = (-10, 0),
    xlabel="x",ylabel="y")
```

## Two-dimensional diffusion and advection {-}

$$
u_t = \phi(u, u_x, u_y, u_{xx}, u_{xy}, u_{yy}), \quad (x,y) \in [a,b] \times [c,d].
$$

- assuming periodic conditions or Dirichlet boundary conditions
- goal: express as an IVP to be solved with Runge-Kutta
    + solution: vectorize the matrix
- **periodic end conditions**: vectorize the initial condition and use ODEProblem()

ex: heat equation

```{julia}
m = 60;  x,Dx,Dxx = FNC.diffper(m,[-1,1]);
n = 25;  y,Dy,Dyy = FNC.diffper(n,[-1,1]);
mtx = f -> [ f(x,y) for x in x, y in y ];
unvec = z -> reshape(z,m,n);
u_init = (x,y)->sin(4*π*x)*exp(cos(π*y));
U0 = mtx(u_init);
M = maximum(abs,U0);

contour(x,y,U0',color=:redsblues,clims=(-M,M),aspect_ratio=1,
    xaxis=("x",(-1,1)),yaxis=("y",(-1,1)),title="Initial condition") 
    
function dudt(u,α,t)
    U = unvec(u)
    Uxx = Dxx*U;  Uyy = U*Dyy'     # 2nd partials
    dUdt = α*(Uxx + Uyy);          # PDE
    return vec(dUdt)
end;

IVP = ODEProblem(dudt,vec(U0),(0,0.2),0.1);
sol = solve(IVP,Rodas4P());

t = 0;
surface(x,y,unvec(sol(t))',color=:redsblues,clims=(-M,M),
    xaxis=(L"x",(-1,1)),yaxis=(L"y",(-1,1)),zlims=(-M,M),
    title=@sprintf("Heat equation, t=%.3f",t),
    dpi=100,colorbar=:none)
```


- **Dirichlet conditions**: boundary conditions correspond to entire rows and columns of the matrix as compared to chapter 11
    + define pack and unpack columns to extend boundary values by zeros

ex: $u_t + u_x = 1 + \epsilon(u_{xx} + u_{yy}), u = 0$ on boundary of $[-1,1]^2$

```{julia}
m = 50; n = 36;
x,Dx,Dxx = FNC.diffcheb(m,[-1,1]);
y,Dy,Dyy = FNC.diffcheb(n,[-1,1]);
U0 = [ (1+y)*(1-x)^4*(1+x)^2*(1-y^4) for x in x, y in y ];
Ex = [zeros(m-1) I(m-1) zeros(m-1)];
Ey = [zeros(n-1) I(n-1) zeros(n-1)];
unvec = u -> reshape(u,m-1,n-1);

pack = U -> vec(Ex*U*Ey');
unpack = w -> Ex'*unvec(w)*Ey;

function dwdt(w,ϵ,t)
    U = unpack(w)
    Ux,Uxx = Dx*U , Dxx*U
    Uyy = U*Dyy'
    dUdt = @. 1 - Ux + ϵ*(Uxx + Uyy)
    return pack(dUdt)
end

IVP = ODEProblem(dwdt,pack(U0),(0.,2),0.05);
w = solve(IVP,Rodas4P());
    

U = unpack(w(t));
surface(x,y,U',layout=(1,2),size=(640,320),
    xlabel=L"x",ylabel=L"y",zaxis=((0,2),L"u(x,y)"),
    color=:viridis,alpha=0.66,clims=(0,2),colorbar=:none,
    title="Advection-diffusion",dpi=100 );
contour!(x,y,U',levels=24,aspect_ratio=1,
    subplot=2,xlabel=L"x",ylabel=L"y",
    color=:viridis,clims=(0,2),colorbar=:none,
    title=@sprintf("t = %.2f",t) )
```

- wave equation demo

## Exercise 13.2.2 {-}

Solve the Allen-Cahn equation $u_t = u(1-u^2) + 0.0001(u_{xx} + u_{yy})$ on the square $[-1,1]^2$ with periodic conditions, taking $u(x,y,0) = \sin(\pi x)\cos(2\pi y)$. Use $m=n=60$ to solve up to $t=4$, and make an animation.

```{julia, eval = F}
m = 60;  x,Dx,Dxx = FNC.diffper(m,[-1,1]);
n = 60;  y,Dy,Dyy = FNC.diffper(n,[-1,1]);
mtx = f -> [ f(x,y) for x in x, y in y ];
unvec = z -> reshape(z,m,n);
u_init = (x,y)->sin(π*x)*cos(2π*y);
U0 = mtx(u_init);
M = maximum(abs,U0);

contour(x,y,U0',color=:redsblues,clims=(-M,M),aspect_ratio=1,
    xaxis=("x",(-1,1)),yaxis=("y",(-1,1)),title="Initial condition") 
    
function dudt(u,α,t)
    U = unvec(u)
    Uxx = Dxx*U;  Uyy = U*Dyy'     # 2nd partials
    dUdt = U.*(1 .- U^2) + α*(Uxx + Uyy);          # PDE
    return vec(dUdt)
end;

IVP = ODEProblem(dudt,vec(U0),(0,4.0),0.0001);
sol = solve(IVP,Rodas4P());

t = 0;
surface(x,y,unvec(sol(t))',color=:redsblues,clims=(-M,M),
    xaxis=(L"x",(-1,1)),yaxis=(L"y",(-1,1)),zlims=(-M,M),
    title=@sprintf("Heat equation, t=%.3f",t),
    dpi=100,colorbar=:none)
```


## Exercise 13.2.5 {-}

From Maxwell’s equations we can find a way to convert the wave equation to a first-order form that uses only first-order derivatives in space:

$$
u_t = c^2(v_y - w_x),\\
v_t = u_y\\
w_t = -u_x,
$$
subject to $u=0$ on the boundary

(a) Show that a solution satisfies $u_t = c^2(u_{xx} + u_{yy})$

$$
v_{ty} = u_{yy}\\
w_{tx} = -u_{xx}
$$

Now what?

(b) Solve with $c=2$ in the rectange $[-3,3] \times [-1,1]$, $u(x,y,0) = \exp(x-x^2)(9-x^2)(1-y^2)$, and $v=w=0$ at $t=0$. Use $m=50$ for x and $n=25$ for y, solve for $0 \leq t \leq 6$, and make an animation.

```{julia}
m = 50;  n = 25;
x,Dx,Dxx = FNC.diffcheb(m,[-3,3]);
y,Dy,Dyy = FNC.diffcheb(n,[-1,1]);
U0 = [ exp(x-x^2)*(9-x^2)*(1-y^2) for x in x, y in y ]
V0 = zeros(size(U0));
```




## Laplace and Poisson equations {-}

The **Poisson equation** in two dimensions is

$$
u_{xx} + u_{yy} = f(x,y),
$$
where $\Delta u = f$ and $\Delta$ is the **Laplacian operator** and sometimes called a *forcing function*. If $f$ is 0, then it is the **Laplace equation**.

- archetype of an **elliptic PDE**
- no time appears
- often represent steady state
- must complement with boundary condition; only consider $u(x,y) = g(x,y)$

## Sylvester equation {-}

- discretized Poisson equation:

$$
\mathbf{D}_{xx}\mathbf{U} + \mathbf{U}\mathbf{D}_{yy}' = \mathbf{F}
$$

- Kronecker product, $\otimes$:

```{julia}
A = [1 2; -2 0]
B = [ 1 10 100; -5 5 3 ]
A_kron_B = [ A[1,1]*B  A[1,2]*B;
             A[2,1]*B  A[2,2]*B ]
kron(A,B)
```

- Poisson as a linear system:

$$
[(\mathbf{I}_y \otimes \mathbf{D}_{xx}) + (\mathbf{D}_{yy} \otimes \mathbf{I}_x)]\text{vec}(\mathbf{U}) = \text{vec}(\mathbf{F})\\
\mathbf{Au} = \mathbf{b}.
$$

- add boundary conditions to their precise values as in collocation strategy
- template: create the linear system, modify it for the boundary conditions, solve it using backslash, and reshape to get a grid function
    + condition number improved by rescaling by the largest of linear coefficients
- walk through `poissonfd`

ex:

```{julia}
f = (x,y) -> x^2 - y + 2;
m,n = 6,5
x,Dx,Dxx = FNC.diffmat2(m,[0,3]);
y,Dy,Dyy = FNC.diffmat2(n,[-1,1]);
unvec = u -> reshape(u,m+1,n+1);

F = [ f(x,y) for x in x, y in y ]
A = kron(I(n+1),Dxx) + kron(Dyy,I(m+1));
b = vec(F);

@show N = length(F);

spy(sparse(A),color=:blues,m=3,
    title="System matrix before boundary conditions")
isboundary = trues(m+1,n+1);
isboundary[2:m,2:n] .= false;
idx = vec(isboundary);

spy(sparse(isboundary),m=3,color=:darkblue,legend=:none,
    title="Boundary points",
    xaxis=("column index",[0,n+2]),yaxis=("row index",[0,m+2]) )

I_N = I(N);
A[idx,:] .= I_N[idx,:];     # Dirichlet conditions

spy(sparse(A),color=:blues,m=3,
    title="System matrix with boundary conditions") 

b[idx] .= 0;                 # Dirichlet values

u = A\b;
U = unvec(u)
```

- Accuracy and efficiency assuming square grid and finite differencing
    + $\mathbf{A}$ has size $N = O(n^2)$ with sparse Cholesky $O(n^4)$
    + $n$ increases $\rightarrow$ lower truncation error, more operations
    + for fixed run time, $T$, convergence is $O(1/\sqrt{T})$ where $n = O(T^{1/4})$
    + for dense Chebyshev spectral discretization, convergence is $O(K^{-T^{1/6}})$ or $O(K^{-n})$ 

## Exercise {-}

## Nonlinear elliptic PDEs {-}

Generally:

$$
\phi(x,y, u, u_x, u_y, u_{xx}, u_{yy}) = 0, \quad (x,y) \in (a,b) \times (c,d).
$$
with boundary condition $u(x,y) = g(x,y)$.

- core: formulate collocation equations at grid points based on discrete approx and solve by quasi-Newton method
- read `elliptic` function
- off-grid evaluation done by global polynomial interpolation:
    + $\mathbf{U} = \text{mtx}(u)$
    + interpolate across column $\mathbf{u}_j$ to get $v_j = p_j(\xi)$
    + then interpolate over $v_j$ and nodes in $y$

Demo: micromechanical deflector

```{julia}
λ = 1.5;
ϕ = (X,Y,U,Ux,Uxx,Uy,Uyy) -> @. Uxx + Uyy - λ/(U+1)^2;
g = (x,y) -> 0; 
u = FNC.elliptic(ϕ,g,15,[0,2.5],8,[0,1]);

x = range(0,2.5,length=100);
y = range(0,1,length=50);
U = [u(x,y) for x in x, y in y];
contourf(x,y,U',color=:viridis,aspect_ratio=1,
    xlabel=L"x",ylabel=L"y",zlabel=L"u(x,y)",
    title="Deflection of a MEMS membrane",
    right_margin=3Plots.mm)      

## check boundary is zero
x = range(0,2.5,length=100);
norm( [u(x,0) - g(x,0) for x in x], Inf )

## truncation error
[ u(x,y) for x in 0.5:0.5:2, y in 0.25:0.25:0.75 ]
u = FNC.elliptic(ϕ,g,25,[0,2.5],14,[0,1]);
[ u(x,y) for x in 0.5:0.5:2, y in 0.25:0.25:0.75 ]
```

## Exercise 13.4.2 {-}

(a) Solve for steady state of

$$
u_t = -u_y - x - 2 + \epsilon(u_{xx} + u_{yy})
$$
for $\epsilon = 1$ in $[-1,1]^2$, subject to a homogeneous Dirichlet boundary condition. Use $m = n = 30$

```{julia}

```


## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
